Accessing biological data with semantic web technologies

### 2013-12-30 article.tex
Wrote up another example, something I actually frequently do

>>>>

A second example combines regular REST or -like services that return RDF with a local SPARQL query.
For example, this is the approach used in Bioclipse to interact with many of the OpenTox
services~\cite{Willighagen2013}, but can be used for the Open PHACTS Linked Data API
too~\cite{Williams2012}. For example, we can get data from an OpenTox data set using
RCurl~\cite{Hardy2010,Lang2013}, in this example data set 112 with 320 compounds,
which originates from the ToxCast project~\cite{Dix2007}:

\begin{footnotesize}
\begin{verbatim}
library(RCurl)
rdfContent = getURL(
  paste(
    "http://apps.ideaconsult.net:8080/ambit2/dataset/112/compounds",
    "media=text/n3",
    sep="?"
  ),
  write=basicTextGatherer()
)
store = fromString.rdf(rdfContent, format="N3")
compounds = sparql.rdf(store,
  paste(
    "PREFIX ot: <http://www.opentox.org/api/1.1#>",
    "SELECT DISTINCT ?compound WHERE {",
    " ?compound a ot:Compound",
    "}"
  )
)
\end{verbatim}
\end{footnotesize}

In fact, we can continue getting further detail about these compound by repeating the same
approach on the compound URIs. Let's assume we are further interested in the SMILES string.
This feature has the URI \url{http://apps.ideaconsult.net:8080/ambit2/feature/21753}. We
add this as the list of features to retrieve:

\begin{footnotesize}
\begin{verbatim}
feature = "http://apps.ideaconsult.net:8080/ambit2/feature/21753"
getOptions = paste(
  paste("feature_uris[]", curlEscape(feature), sep="="),
  paste("media", curlEscape("application/rdf+xml"),sep="="),
  sep="&"
)
\end{verbatim}
\end{footnotesize}

We can then use the R apply() method to retrieve the SMILES for each structure with:

\begin{footnotesize}
\begin{verbatim}
result = apply(compounds, MARGIN=1, function(x) {
  compound = x["compound"];
  compoundURI = paste(compound, getOptions, sep="?")
  print(paste("Downloading",compoundURI))
  cmpdContent = getURL(
    compoundURI, write=basicTextGatherer()
  )
  fromString.rdf(cmpdContent, format="RDF/XML", appendTo=store)
})
\end{verbatim}
\end{footnotesize}

Finally, we use SPARQL again to list all the SMILES strings for the compounds:

\begin{footnotesize}
\begin{verbatim}
query =  paste(
  "PREFIX ot: <http://www.opentox.org/api/1.1#> ",
  "SELECT DISTINCT ?compound ?value WHERE {",
  " ?s ?o [ ot:feature <", feature, "> ;",
  "   ot:value ?value ",
  " ] ;",
  "   ot:compound ?compound . ",
  "}",
  sep=""    
)
compoundSMILESes = sparql.rdf(store, query)
\end{verbatim}
\end{footnotesize}

This results in a matrix with compound URIs in one column and SMILES strings in another, ready for
use by other R packages.
